Eco-evolutionary dynamics of finite populations from first principles

### 2023-03-14 ind_files/01_Introduction.tex
Change filenames to be more sensible

\section{Idealization and generality in ecology}\label{idealization}
The push and pull between the search for general patterns and the specification of minutiae has a long and torturous history in ecology~\citep{kingsland_modeling_1985} and evolution~\citep{provine_origins_2001}. One of the first mathematical idealizations in ecology came in the form of the logistic equation formulated by Pearl and Verhulst, and shortly thereafter, equations for the populations of interacting predator and prey species, put forth independently by Lotka and Volterra. These models were immediately controversial, and for good reason: Many ecologists felt that they were overly idealistic and neglected many important truths about real biological populations~\citep{kingsland_modeling_1985}. These models were nevertheless quite good at predicting the patterns of such populations, and have proved valuable to the field of population ecology. Today, these models are viewed as `classical' and are regularly used even by hard-line empiricists, not because we believe them to be true in all their gory biological details, but because we recognize that they can be \emph{useful} despite being blatantly false generalizations. This is a single instance of a more general philosophical idea concerning the goals and ideals of science. Some philosophers of science have recently argued that ``the epistemic goal of science is not truth, but understanding''~\citep{potochnik_idealization_2018}. The fact that idealization is part and parcel of science is clear if one looks at the actual practice, be it theorists making unrealistic assumptions on paper to model specific phenomena or experimentalists creating artificially controlled conditions in the laboratory to test specific hypotheses~\citep{zuk_models_2018}. Relatively simple models of complex eco-evolutionary processes are therefore desirable as a way to shine light on these phenomena.

To reiterate once more, the goal of such simple models and organizing frameworks is not truth (whatever that is), predictive power (as with models in physics), or detailed description (as with detailed individual-based simulations, very flexible statistical models, or machine learning), but \emph{understanding}. Since the world is complicated and humans are limited, such understanding inevitably comes at the cost of other desirable qualities such as the ability to make precise quantitative predictions. It is important to remember at the outset that the models I speak about in this thesis may seem to be too general and may seem to ignore many biological details. This is \textit{by design}, in pursuit of general insight over precise quantitative prediction.

A general organization like this will often not be perfect or all-encompassing. As Robert MacArthur once remarked, ``general events are only seen by ecologists with rather blurred vision. The very sharp-sighted always find discrepancies and are able to see that there is no generality, only a spectrum of special cases”\citep{kingsland_modeling_1985}. However, formulating such (blurry-eyed) general organizing frameworks can be greatly beneficial as an aid to thinking. This is perhaps best illustrated by the success of statistical mechanics in physics - Statistical mechanics was essentially born from the idea that statistical regularities of systems can emerge without the need for knowing the excruciating details of every single moving part, and indeed, this sort of explicitly `blurry-eyed' thinking was shown to be able to recover the phenomenological laws of thermodynamics from first principles. In biology, arguably the greatest such general organizing framework was the idea of evolution by natural selection as synthesized from myriad detailed observations by Charles Darwin. Theoretical population genetics has also had a long-standing tradition in building general organizing frameworks that `abstract away' some biological details but are `general' in the sense of capturing broad features of biological systems (as opposed to the mathematician's notion of `general' in sense of always being true). In a similar spirit, Vellend has recently argued that conceptual synthesis in community ecology requires
``shifting the emphasis away from an organizational structure based on the useful lines of
inquiry carved out by researchers, to one based on the fundamental processes that underlie community dynamics and patterns''~\citep{vellend_theory_2016}. Vellend's assertion is based on the
fact that population genetics has managed to come up  with reasonably comprehensive theory due to its focus on the abstract `high-level’ processes of selection, mutation, drift, and
gene flow (see section below) instead of the myriad `low-level' processes that may be responsible for generating them. In contrast, he believes that practitioners of community ecology often focus on
specific `low-level' processes such as predation rate, limiting resources (R\textsuperscript{*}), storage effects,
priority effects, senescence, and niche partitioning, leading to a plethora of models (see Table
5.1 in~\cite{vellend_theory_2016} for an in-exhaustive list of 24 such models) and the conclusion that
community ecology `is a mess'. Vellend instead proposes organizing ecological models according to
 the `high-level’ processes of selection, ecological drift (demographic stochasticity), dispersal,
and speciation, in analogy with population genetics.

\section{A (very, very) brief history of high-level modelling frameworks in population biology}

@@

One of the general guiding principles of much of this mathematization has been the assumption that incorporating the reality of finite population sizes into models leads to no major qualitative differences in behavior, only `adding noise' or `blurring out' the predictions of simpler infinite population models~\citep{page_unifying_2002}. Consequently, several major theoretical frameworks in the field, such as adaptive dynamics, are explicitly formulated in deterministic terms at the infinite population size limit. However, this assumption is largely unjustified, and since populations in the real world are finite and stochastic, checking whether stochastic models differ from their deterministic analogs is vital to furthering our understanding of the fundamentals of population biology~\citep{hastings_transients_2004, coulson_skeletons_2004, shoemaker_integrating_2020}.Today, we increasingly recognize that incorporating the finite and stochastic nature of the real world routinely has much stronger consequences than simply `adding noise' to deterministic expectations~\citep{boettiger_noise_2018}, with important consequences for both ecological~\citep{schreiber_does_2022} and evolutionary~\citep{delong_stochasticity_2023} theory. In ecology and evolution, stochastic models often do not exhibit phenomena that occur in deterministic models~\citep{proulx_what_2005, johansson_will_2006, claessen_delayed_2007,  wakano_evolutionary_2013, debarre_evolutionary_2016, johnson_two-dimensional_2021}, exhibit phenomena that do not occur in deterministic models~\citep{rogers_demographic_2012, rogers_spontaneous_2012, rogers_modes_2015, veller_drift-induced_2017, delong_stochasticity_2023}, and can even completely reverse the predictions of deterministic models~\citep{houchmandzadeh_selection_2012,houchmandzadeh_fluctuation_2015,constable_demographic_2016,mcleod_social_2019}. Studies of neutral or near-neutral dynamics in population and quantitative genetics usually do take stochasticity seriously, explicitly modeling finite populations that follow stochastic dynamics. Unfortunately, the classic or standard stochastic models in both population genetics~\citep{fisher_genetical_1930,wright_evolution_1931, moran_random_1958, kimura_diffusion_1964} and quantitative genetics~\citep{crow_introduction_1970, lande_natural_1976} typically assume a fixed total population size and their validity is therefore rather restrictive since population sizes routinely fluctuate in the real world. Importantly, the Price equation itself is usually formulated in a deterministic, dynamically-insufficient manner (but see~\cite{rice_stochastic_2008} for a stochastic formulation of the Price equation). Since real-life populations are stochastic, finite, and of non-constant population size, it is thus imperative that we develop a theoretical framework that can handle such systems directly, instead of only working with deterministic, infinite-population approximations.

>>>>

\section{Idealization and generality}\label{idealization}
The push and pull between the search for general patterns and the specification of minutiae has a long and torturous history in ecology~\citep{kingsland_modeling_1985} and evolution~\citep{provine_origins_2001}. The fact that idealization is part and parcel of science is clear if one looks at the actual practice, be it theorists making unrealistic assumptions on paper to model specific phenomena or experimentalists creating artificially controlled conditions in the laboratory to test specific hypotheses~\citep{zuk_models_2018}. Indeed, some philosophers of science recently argued that ``the epistemic goal of science is not truth, but understanding''~\citep{potochnik_idealization_2018}, an idea generally echoed by practicing scientists~\citep{levins_strategy_1966,zuk_models_2018,grainger_empiricists_2022}.  In other words, since the world is complicated and humans are limited, general understanding inevitably comes at the cost of other desirable qualities such as the ability to make precise quantitative predictions. This is especially true for complex phenomena such as those that are the domain of ecology and evolution, where we are often not even aware of all the factors that are at play or how they interact. It is important to remember at the outset that the models I speak about in this thesis may seem to be too abstract/general and may seem to ignore many relevant biological details. This is \textit{by design}, in pursuit of general insight over precise quantitative prediction.

An approach like this will often not be perfect or all-encompassing. As Robert MacArthur once remarked, ``general events are only seen by ecologists with rather blurred vision. The very sharp-sighted always find discrepancies and are able to see that there is no generality, only a spectrum of special cases”\citep{kingsland_modeling_1985}. However, formulating and studying such generalizations can be greatly beneficial as an aid to thinking, sometimes by the very virtue of the fact that it can be done through `blurry eyed' thinking that only looks for general broad-brush regularities\footnote{It bears noting that when a biologist says something is `general', they typically mean that it occurs in the majority of the relevant cases studied. In contrast, when a mathematician says something is `general', they mean that it is \emph{always} true. Thus, an event that occurs in 70\% of cases is `generally true' to biologists but `not true in general' to mathematicians. In this thesis I write as a biologist.}. This is perhaps best illustrated by the success of statistical mechanics in physics - Statistical mechanics was essentially born from the idea that statistical regularities of systems can emerge without the need for knowing the excruciating details of every single moving part, and indeed, starting from first principles, this sort of explicitly `blurry-eyed' thinking was shown to be able to recover the phenomenological laws of thermodynamics as \emph{statistical laws}. In biology, arguably the greatest such general organizing framework was the idea of evolution by natural selection as synthesized from myriad detailed observations by Charles Darwin. Theoretical population genetics has also had a long-standing tradition in building general organizing frameworks that `abstract away' some biological specificities in favor of a small number of `fundamental' forces, and here, the best analog of statistical mechanics is the Price equation, which was able to recover several known formalisms of evolution from first principles (see section \ref{sec_history} below). In a similar spirit, Vellend has recently argued that conceptual synthesis in community ecology requires
``shifting the emphasis away from an organizational structure based on the useful lines of
inquiry carved out by researchers, to one based on the fundamental processes that underlie community dynamics and patterns''~\citep{vellend_theory_2016}. Vellend's assertion is based on the
fact that population genetics has managed to come up  with reasonably comprehensive theory due to its focus on the abstract `high-level’ processes of selection, mutation, drift, and
gene flow (see section below) instead of the myriad `low-level' processes that may be responsible for generating them. In contrast, he believes that practitioners of community ecology often focus on
specific `low-level' processes such as predation rate, limiting resources (R\textsuperscript{*}), storage effects,
priority effects, senescence, and niche partitioning, leading to a plethora of models (see Table
5.1 in~\cite{vellend_theory_2016} for an in-exhaustive list of 24 such models) and the conclusion that
community ecology `is a mess'. Vellend instead proposes organizing ecological models according to
 the `high-level’ processes of selection, ecological drift (demographic stochasticity), dispersal,
and speciation.

\section{A (very, very) brief history of high-level modelling frameworks in population biology}\label{sec_history}

@@

One of the general guiding principles of much of this mathematization has been the assumption that incorporating the reality of finite population sizes into models leads to no major qualitative differences in behavior, only `adding noise' or `blurring out' the predictions of simpler infinite population models~\citep{page_unifying_2002}. Consequently, several major theoretical frameworks in the field, such as adaptive dynamics, are explicitly formulated in deterministic terms at the infinite population size limit. However, this assumption is largely unjustified, and since populations in the real world are finite and stochastic, checking whether stochastic models differ from their deterministic analogs is vital to furthering our understanding of the fundamentals of population biology~\citep{hastings_transients_2004, coulson_skeletons_2004, shoemaker_integrating_2020}. Today, we increasingly recognize that incorporating the finite and stochastic nature of the real world routinely has much stronger consequences than simply `adding noise' to deterministic expectations~\citep{boettiger_noise_2018}, with important consequences for both ecological~\citep{schreiber_does_2022} and evolutionary~\citep{delong_stochasticity_2023} theory. In ecology and evolution, stochastic models often do not exhibit phenomena that occur in deterministic models~\citep{proulx_what_2005, johansson_will_2006, claessen_delayed_2007,  wakano_evolutionary_2013, debarre_evolutionary_2016, johnson_two-dimensional_2021}, exhibit phenomena that do not occur in deterministic models~\citep{rogers_demographic_2012, rogers_spontaneous_2012, rogers_modes_2015, veller_drift-induced_2017, delong_stochasticity_2023}, and can even completely reverse the predictions of deterministic models~\citep{houchmandzadeh_selection_2012,houchmandzadeh_fluctuation_2015,constable_demographic_2016,mcleod_social_2019}. Studies of neutral or near-neutral dynamics in population and quantitative genetics usually do take stochasticity seriously, explicitly modeling finite populations that follow stochastic dynamics. Unfortunately, the classic or standard stochastic models in both population genetics~\citep{fisher_genetical_1930,wright_evolution_1931, moran_random_1958, kimura_diffusion_1964} and quantitative genetics~\citep{crow_introduction_1970, lande_natural_1976} typically assume a fixed total population size and their validity is therefore rather restrictive since population sizes routinely fluctuate in the real world. Importantly, the Price equation itself is usually formulated in a deterministic, dynamically-insufficient manner (but see~\cite{rice_stochastic_2008} for a stochastic formulation of the Price equation). Since real-life populations are stochastic, finite, and of non-constant population size, it is thus imperative that we develop a theoretical framework that can handle such systems directly, instead of only working with deterministic, infinite-population approximations.
